Water discharge has a variable impact on sediment transport capacity in subglacial systems

### 2023-08-25 Delaneyetal2023b.tex
small changes

\item Water discharge may represent processes including proglacial sediment mobilization and sediment access driving variations in sediment discharge.
\end{itemize}

@@

Second, to characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels further, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations  are culled if their mean subglacial water velocity over the season lies outside of the range between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies outside of the range $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with some measurements \citep[e.g.]{werder2010b,magnusson2012,chandler2013}.
The standard deviation of an output variable over the course of the season is used to quantify the variability of thats output.

@@

The strong link between sediment discharge capacity and open channel flow conditions could be manifested ~ $20$\,\unit{km} downstream of the Leverett measurement site, where a strong correlation persists between sediment plume size and water discharge into the Kangerlussuaq fjord \citep{mcgrath2010}.
This lies in contrast to other marine-terminating glacier fjords in Greenland where the subglacial channels are pressurized by ocean water at the ice front \citep[e.g.][]{how2017}. 
In these catchments, a less consistent relationships may persist between water discharge, or melt extend, and sediment plume size \citep{tedstone2012}, possibly due to the variable link between sediment transport capacity and water discharge in subglacial conditions.

Second, the sediment discharge records could represent factors such as increased sediment accessibility when melt extends up glacier, thereby concurrently increasing sediment discharge with water discharge \citep[e.g.][]{delaney2020}.
In this case, the sediment transport records presented here may represent supply-limited sediment transport, less dependent on sediment transport capacity.
Glaciers could well trend toward a supply-limited regime, because the threshold of motion for sediment can be reached more frequently, and across a range of water discharges, compared to subaerial systems.
In a supply-limited system, sediment discharge would also scale with sediment production from glacier quarrying and abrasion, along with meltwater access \citep[e.g.][]{herman2015,ugelvig2018,delaney2023}.
 Sediment exhaustion through this process may also explain the stronger dependence of sediment discharge from the Greenland Ice Sheet on basal shear stress, a proxy for bedrock erosion, rather than glacier melt \citep{overeem2017}.
Note, however, that the absence of lags between sediment and water discharge may suggest that sediment is mobilized close to the measurement stations, as opposed to distal areas of the glacier bed as in the supply-limited regime \citep[Figure~\ref{fig:model_outs}\,a and e][]{williams1989}.

Regardless of the increased performance of water discharge compared to subglacial sediment transport capacity in the observations, parameterization outputs suggest that additional processes and erosional hiatuses may further complicate signals of sediment discharge from glaciers in response to climate, compared to subaerial systems \citep{jansson2005,ganti2016}.
Records of sediment discharge have been used to establish the relationship, or lack thereof, between sediment discharge and climate in glacial systems \citep[e.g.][]{koppes2009a,willenbring2016,mariotti2021}.
The variable relationship between water discharge and subglacial sediment transport capacity presented here mostly applies to the daily- to weekly-time scales controlling the size of subglacial channels (Figure~\ref{fig:multi_run}).
Yet, identifying climatic signals in sediment transport from transport-limited glaciers may require higher thresholds of water discharge compared to subaerial systems \citep[Figure~\ref{fig:Qw_vari}][]{tofelde2021}.
This higher threshold of perturbation by water discharge, which impacts sediment transport capacity, might compound with the stochastic nature of erosion and deposition in fluvial environments \citep{castletort2003,jerolmack2010,romans2016}.

In addition to water discharge, the glacier's sediment transport capacity reacts to the ice thickness, controlling the closure of the channel, and the surface slope of the glacier, controlling water velocity \citep[Section~\ref{sect:sub_mode}; ][]{rothlisberger1972,shreve1972,stevens2022}.
Conversely, sediment transport in most subaerial systems typically increases with water discharge and hydraulic gradient, and the hydraulic gradient of most subaerial systems likely remains stable over time scales ranging from years to millennia \citep[Section~\ref{sect:fluv}; e.g.][]{muller1968,whipple1999,wong2006,wickert2019}. 
% Furthermore, if no sediment is available and the glacier is in a supply-limited state, often assumed in glacier landscape evolution models, then the sediment discharge record also represents additional processes of bedrock erosion from  water pressure variations and sliding  \citep{iverson2012,herman2015}, which are also complex.

\ian{Comeback here and link with paragraph above better.}

@@

Implications of the findings here include:
\begin{enumerate}
\item the inconsistent relationship between water discharge and sediment transport capacity underneath glaciers, causing increased variability in sediment transport capacity compared to river systems;
\item the timing of peak sediment transport capacity below glaciers generally coincides with large rates of increasing in water discharge, as opposed to the maximum quantity of water discharge;
\item the possibility for a consistent relationship between water and sediment discharge to represent processes such as subglacial sediment access, or open channel processes near the glacier margin, instead of subglacial sediment transport capacity.

>>>>

\item In interpreting sediment transport records, water discharge may represent processes including proglacial sediment mobilization and sediment access driving variations in sediment discharge.
\end{itemize}

@@

Second, to characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels further, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations  are culled if their mean subglacial water velocity over the season lies outside of the range between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies outside of the range $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with some measurements \citep[e.g.]{werder2010b,magnusson2012,chandler2013}.
The standard deviation of an output variable over the course of the season is used to quantify the variability of that output.

@@

The strong link between sediment discharge capacity and open channel flow conditions could be manifested ~ $20$\,\unit{km} downstream of the Leverett measurement site, where a strong correlation persists between sediment plume size and water discharge into the Kangerlussuaq fjord \citep{mcgrath2010}.
This lies in contrast to other marine-terminating glacier fjords in Greenland where the subglacial channels are pressurized by ocean water at the ice front \citep[e.g.][]{how2017}. 
In these catchments, a less consistent relationships may persist between water discharge or melt extend and sediment plume size \citep{tedstone2012}, possibly due to the variable relationship between subglacial sediment transport capacity and water discharge.

Second, the sediment discharge records could represent factors such as increased sediment accessibility when melt extends up glacier, thereby concurrently increasing sediment discharge with water discharge \citep[e.g.][]{delaney2020}.
In this case, the sediment transport records presented here may represent supply-limited sediment transport, less dependent on sediment transport capacity.
Glaciers could well trend toward a supply-limited regime, because the threshold of motion for sediment can be reached more frequently, and across a range of water discharges, compared to subaerial systems (Figure~\ref{fig:Qw_vari}).
In a supply-limited system, sediment discharge would also scale with sediment production from glacier quarrying and abrasion, along with meltwater access \citep[e.g.][]{herman2015,ugelvig2018,delaney2023}.
 Sediment exhaustion through this process may also explain the stronger dependence of sediment discharge from the Greenland Ice Sheet on basal shear stress, a proxy for bedrock erosion, rather than glacier melt \citep{overeem2017}.
Note, however, that the absence of lags between sediment and water discharge may suggest that sediment is mobilized close to the measurement stations, as opposed to distal areas of the glacier bed as in the supply-limited regime \citep[Figure~\ref{fig:model_outs}\, a and e; ][]{williams1989}.

Regardless of the increased performance of water discharge compared to subglacial sediment transport capacity in the observations, parameterization outputs suggest that additional processes and erosional hiatuses may further complicate signals of sediment discharge from glaciers in response to climate, compared to subaerial systems \citep{jansson2005,ganti2016}.
Records of sediment discharge have been used to establish the relationship, or lack thereof, between sediment discharge and climate in glacial systems \citep[e.g.][]{koppes2009a,willenbring2016,mariotti2021}.
The variable relationship between water discharge and subglacial sediment transport capacity presented here mostly applies to the daily- to weekly-time scales controlling the size of subglacial channels (Figure~\ref{fig:multi_run}).
Yet, identifying climatic signals in sediment transport from the transport-limited component of glacier systems may require higher thresholds of water discharge compared to subaerial systems \citep[Figure~\ref{fig:Qw_vari}; ][]{tofelde2021}.
This occurs as the variable relationship between water discharge and sediment transport capacity
may compound  the already stochastic nature of erosion and deposition in fluvial environments \citep{castletort2003,jerolmack2010}.

Over time scales longer than days to weeks, the glacier's sediment transport capacity reacts to the ice thickness, controlling the channel closure rate, and the surface slope of the glacier \citep[Figure~\ref{fig:multi_run}, ~ Section~\ref{sect:sub_mode}; ][]{rothlisberger1972,shreve1972,stevens2022}.
If the glacier is in a supply-limited state, often assumed in landscape evolution models, then the sediment discharge record also represents additional processes of sediment availability and bedrock erosion from  water pressure variations and sliding  \citep[e.g.][]{iverson2012,herman2015,delaney2023}.
This multitude of processes lies in contrast to most subaerial systems, where transport capacity typically increases with water discharge and hydraulic gradient, and the hydraulic gradient generally remains stable over time scales ranging from years to millennia \citep[Section~\ref{sect:fluv}; e.g.][]{muller1968,whipple1999,wickert2019}.
The stronger link between sediment export and hydrological forcing in subaerial channels means that variations in sediment transport records can be more easily attributed to climatic conditions than in  subglacial systems

@@

Implications of the findings here include:
\begin{enumerate}
\item the inconsistent relationship between water discharge and sediment transport capacity underneath glaciers, causing increased variability in sediment transport capacity compared to river systems;
\item the timing of peak sediment transport capacity below glaciers generally coincides with large rates of increasing in water discharge, as opposed to the maximum quantity of water discharge;
\item the possibility for a consistent relationship between water and sediment discharge to represent subglacial sediment access or open channel hydraulics near the glacier margin, instead of subglacial sediment transport capacity.
